Measuring the double-drift illusion with hand tracking

### 2021-06-22 Manuscript.Rmd
working on manuscript

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (middle), we can see that the reset points vary more in when they occur, than how far they are from the real path of the stimulus. That is, predicting reset points from a single Y coordinate (time limit) would leave larger residuals than predicting reset points from a single X coordinate (or spatial limit). This already indicates that a spatial limit explains more of the variance in reset points than a temporal limit.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models. **Left:** Distance of reset from true path predicted from a single temporal limit. **Middle:** Time of reset predicted from a single spatial limit. **Right:** Both spatial offset and time of resets predicted from a temporal and a spatial limit."}
#plotResetPoints()
plotModels()
```

We test this with three simple models, based on straighforward trigonometry. First, we calculate the slope of a line ($s$, time/illusory offset) through the origin and a sample halfway to the reset point, indicating the strength of the illusion. If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:

[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_y / s$

And a spatial limit ($L_x$) would predict time until the first reset, $y$:

[3]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$

While resets relying on both limits would use these two equations:

[4a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_x + L_y / s$

[4b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s + L_y$

Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope from the start of the trajectory to the halfway point. If we fit these simple models by minimizing the mean square error (errors are the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.82 _s_ (Eq. 2; AIC=-20.3) and a spatial limit of ~2.46 _cm_ (Eq. 3; AIC=-28.4), as illustrated in Figure 6 (left and middle). This is close to, but not the same as, the mean values of the respective coordinates of reset points (Fig. 5, middle). The model with two limits uses a temporal limit of 0.55 _s_ and a spatial limit of 1.89 _cm_ (Eq. 4a,b; AIC=-28.8). The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.016, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.87). This indicates that a spatial limit explains most of the data.

>>>>

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (middle), we can see that the reset points vary more along the y-axis, than along the x-axis. That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.

```{r fig5, echo=FALSE, message=FALSE, fig.width=7, fig.height=5.5, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models. Model limit shown here are given by weighted bootstrapping from all participants, lines are median parameter values, shaded areas the 95% confidence interval of the limits. Gray dots are the reset points from all participants. **Left:** Distance of reset from true path as a single spatial limit. **Right:** Reset coordinates predicted from a single temporal limit (at two speeds)."}
#plotResetPoints()
#plotModels()
plotBootstrappedFits()
```

We test this with a series of simple models, based on straighforward trigonometry. First, we calculate the angle of a line through the origin and a sample on the trajectory halfway to the reset point (with coordinates $X$ and $Y$), as the strength of the illusion ($a = atan2(X,Y)$ expressing the angular deviation the line $x=0$ which is the real path of the gabor). This can then be used to calculate a slope ($s = \frac{sin(a)}{cos(a)}$).

If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would predict the y-coordinate of the first reset, $y$:

[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$

And a temporal limit ($L_t$) would predict both coordinates of the reset, $x$ and $y$:

[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$
[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$

Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus. If we fit the two simple models for each participant separately, by minimizing the mean square error (error as absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit. We can also combine limits to see if they are complementary. Here, we see that for 4/9 participants (same for MSE and AIC) the best fit boils down to a spatial limit as the other limit is set to 0. Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit. This indicates that in this data, a spatial limit explains the reset points best. However even the spatial limit does not explain the data very well.

To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points. Both models based on a spatial limit (eq. 2) and on a temporal limit (eq. 3) as well as a model combining these two were fit 1000 times. In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models. That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret. Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5. Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, most used a normal distribution (6/9) and the rest an exponential distribution (3/9). All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.
